Analysis of competitive surfing tournaments with generalized Bradley-Terry

### 2021-08-26 inst/surfing_jsa.tex
tweaks and spellcheck

  \author{Driver, Timothy${}^{1,2}$ \and
    Hankin, Robin K. S.${}^{2,3}$ \\
    \texttt{hankin.robin@gmail.com}\\(corresponding author)}

  \date{${}^1$ {\bf Tim what do you want here you can put anything}\\
    ${}^{1,2}$ Auckland University of Technology
    Wellesley Street, Auckland, New Zealand\\
    +64-9-9219999x5106\\
    ${}^3$ Corresponding author}

@@

\subsubsection*{Abstract}
   Here we analyse competitive surfing, specifically the 2018 World Surf
League, using formal statistical methods.  We use generalized
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Specifically, we quantify the dominance of
the top competitors using likelihood techniques, and go on to study
the ``Brazilian storm'' phenomenon using reified entities in two ways.
Firstly we assess the supposed Brazilian preference for point and
beach break beaches; and secondly we consider tournament results from
the perspective of tournament theory and test for competitors
modifying their strategy in the presence of compatriot rivals.  We
quantify the evidence for these commonly assumed features of
contemporary competitive surfing and suggest further avenues of
research.
\newpage

@@

The typical format is as follows.  Up to four surfers---the
competitors---are in the water simultaneously, watched by up to five
judges.  Subject to surfing etiquette, competitors are free to catch a
wave at their discretion within the heat's time window, typically
20-30 minutes.  The judges rate each wave ride and award points to the
surfer based on a range of subjective criteria such as stylish
execution and transition of manoeuvres, but also credit elements such
as the difficulty and novelty of the performance.  The top score and
bottom score among the five judges are removed and the remaining three
judges' scores are averaged to give the final score for the surfer for
that wave.  Each surfer's aggregate score is given by the average of
the two highest-scoring waves.  The winner is the surfer with the
highest aggregate score.

@@

The Bradley-Terry model \citep{bradley1954} assigns non-negative strengths
$p_1,\ldots, p_n$ to each of $n$ competitors in such a way that the
probability of $i$ beating $j\neq i$ in pairwise competition is
$\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that $\sum
p_i=1$.  Further, we use a generalization due to \citep{luce1959}, in which
the probability of competitor~$i$ winning in a field of $\left\lbrace
1,\ldots, n\right\rbrace$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
that there is information in the whole of the finishing order, and not
just the first across the line, we can follow \citet{plackett1975} and
consider the runner-up to be the winner among the remaining
competitors, and so on down the finishing order. Without loss of
generality, if the order of finishing were $1,2,3,4,5$, then a
suitable Plackett-Luce likelihood function would be

@@

Table~\ref{resultstable} shows the official results table from the
tour.  We see that Ferreira places first with 59740 points, Medina
second with 56475, and so on.  The overall ranking of the surfers is
determined by summing the season's points.  However, this system
suffers from the defect that the points system itself is intrinsically
arbitrary: the details of the points system does not affect the
competitors' behaviour but can change the overall ranking; this would
suggest that more formal statistical methods are neede for analysis of
such datasets.  The Plackett-Luce likelihood functions are not subject
to such a phenomenon and furnish an objective and coherent method for
ranking competitors.  Further, likelihood-based methods offer one
advantage over points-based systems: the ability to conduct
statistically rigorous tests of meaningful nulls.

@@

Table~\ref{rawresults} shows the first few and last few entries in the
dataset we use.  The first line shows that, at one of the heats in
{\tt wct01}, Colapinto came first, Bailey second, and Wright came
third.  The last line shows that, at {\tt wct11}, Ferreira came first
and Medina second.  We can convert this dataset into a Plackett-Luce
likelihood function but first have to remove the wildcards and also
player Vries whose maximum likelihood strength is zero.  For the first
line we would have Plackett-Luce likelihood function as

@@

[observe that there is no requirement here for the strengths to have
unit sum].  The entire dataset has a likelihood function that
incorporates all 477 lines, but it includes:

@@

Paired comparisons are particularly favourable to Bradley-Terry
because the likelihood function is exponential family and as such is
amenable to analysis via general linear models~\citep{turner2012}.
However, this is not applicable when more than two entities are
compared; and indeed here we see comparisons between three surfers
and, in the case of Jeep Surf Freshwater Pro (wct08), many more.  It
is natural to ask how much information is contained in the
observations where more than two surfers are in the water.

@@

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\citep{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \citep{edwards1992}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result actually reflects genuine
differences in abilities, rather than the result of random chance.
Alternatively, we may view the dataset as being sufficiently
informative to reveal actual differences in surfing ability.  Observe
that the competitors' points (that is, the final column of
table~\ref{resultstable}) cannot be used for this type of
probabalistic analysis: the points awarded are arbitrary and changing
the points system would change the points scored but not affect the
competitors' behaviour.

@@

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact that six out of eight querterfinalists at the 2011 Nike
Lowers Pro were Brazilian; it rose in public conciousness following
the victories of Medina and de Souza in the World Surf League
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \citep{douglas2020}.

@@

\subsection{Brazilian wave type prefereneces}

@@

We now assess the null that Brazilians perform equally well on all
types of wave types.  To that end, we create a reified surfer,
``Brazilian wavetype", whose strength helps all Brazilian surfers
when surfing on their (putative) preferred wave type, an analogue of
the ``home ground monster" who helps the home team as per
equation~\ref{home_ground_advantage}.  Here we modify the likelihood
function only when a single Brazilian is in the water, thus avoiding
possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performce is better on their preferred wave types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
wavetype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_wavetype_support}).

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\citep{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\citep{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic team would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  Such hypotheses are difficult to test with conventional
techniques as competitors's strengths themselves are nuisance
parameters~\citep{basu1977}.

@@

We analysed results from the 2018 World Surf League using reified
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Reified Bradley-Terry allows the use of
formal statistical methods in ways that analysing points scored does
not.  For example, we can be confident that the results reflect
differences in skill (rather than random variation); there is no
statistical difference in strength between the top two officially
ranked competitors, Medina and Ferreira; but we may reject the
hypothesis that Medina and Wilson are equal in strength.

@@

\section*{Tables}

\begin{table}[h]
\begin{tabular}{llllllllllllll}
competitor  & nationality&wct01  &wct02  &wct03  & \ldots &wct09&wct10&wct11&points\\ \hline
 Ferreira   &   BRA      &   1   &   5   &   17  & \ldots      &  2   &   1   &   1   &   59740\\
 Medina     &   BRA      &   5   &   5   &   17  & \ldots      &  9   &   9   &   2   &   56475\\
 Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots      &  9   &   2   &   17  &   49985\\
 Toledo     &   BRA      &   9   &   2   &   5   & \ldots      &  17  &   5   &   17  &   49145\\
 Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655
\end{tabular}
\caption{WCT tour results, conventional league table\label{resultstable}}
\end{table}

\begin{table}[h]
\begin{verbatim}
wct01 Colapinto Bailey Wright
wct01 Freestone Lau Smith
wct01 Dora Ferreira Slater
wct01 Duru Toledo Ibelli
wct01 Moniz Heazlewood Wilson
...
wct11 Medina Florence
wct11 Colapinto Bourez
wct11 Ferreira Slater
wct11 Medina Colapinto
wct11 Ferreira Medina
\end{verbatim}
\caption{Extract from observations: raw results\label{rawresults} from
  WCT 2018}
\end{table}

@@

\clearpage
\newpage
\section*{Figures}

>>>>

  \author{Driver, Timothy C.${}^{1}$ \and
    Hankin, Robin K. S.${}^{1,2}$}

  \date{${}^{1}$ Auckland University of Technology\\
    Wellesley Street, Auckland, New Zealand\\
    +64-9-9219999x5106\\
    ${}^2$ Corresponding author\\
    \texttt{hankin.robin@gmail.com}
  }

  % Tim:   ORCID 0000-0002-3184-2118
  % Robin: ORCID 0000-0001-5982-0415

@@

\subsubsection*{Abstract}
Here we analyse competitive surfing, specifically the 2019 World Surf
League, using formal statistical methods.  We use generalized
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Specifically, we quantify the dominance of
the top competitors using likelihood techniques, and go on to study
the ``Brazilian storm'' phenomenon using reified entities in two ways.
Firstly we assess the supposed Brazilian preference for beach break
and point break wave types; and secondly we consider results from the
perspective of tournament theory and test for competitors modifying
their strategy in the presence of compatriot rivals.  We quantify the
evidence for these commonly assumed features of contemporary
competitive surfing and suggest further avenues of research.
\newpage

@@

The typical format is as follows.  Up to four surfers---the
competitors---are in the water simultaneously, watched by up to five
judges.  Subject to surfing etiquette, competitors are free to catch a
wave at their discretion within the heat's time window.  The judges
rate each wave ride and award points to the surfer based on a range of
subjective criteria such as stylish execution and transition of
manoeuvres, but also credit elements such as the difficulty and
novelty of the performance.  The top score and bottom score among the
five judges are removed and the remaining three judges' scores are
averaged to give the final score for the surfer for that wave.  Each
surfer's aggregate score is given by the average of the two
highest-scoring waves.  The winner is the surfer with the highest
aggregate score.

@@

The Bradley-Terry model \citep{bradley1954} assigns non-negative
strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
that the probability of $i$ beating $j\neq i$ in pairwise competition
is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
$\sum p_i=1$.  Further, we use a generalization due to
\citep{luce1959}, in which the probability of competitor~$i$ winning
in a field of $\left\lbrace 1,\ldots, n\right\rbrace$ is
$\frac{p_i}{p_1+\cdots +p_n}$.  Noting that there is information in
the whole of the finishing order, and not just the first across the
line, we can follow \citet{plackett1975} and consider the runner-up to
be the winner among the remaining competitors, and so on down the
finishing order. Without loss of generality, if the order of finishing
were $1,2,3,4,5$ [we write $1\succ 2\succ 3\succ 4\succ5$], then a
suitable Plackett-Luce likelihood function would be

@@

Table~\ref{resultstable} shows the official results table from the
tour.  We see that Ferreira places first with 59740 points, Medina
second with 56475, and so on.  The overall ranking of the surfers is
determined by summing the season's points.  However, this system
suffers from the defect that the points system itself is intrinsically
arbitrary: the details of the points system does not affect the
competitors' behaviour but can change the overall ranking; this would
suggest that more formal statistical methods are needed for analysis of
such datasets.  The Plackett-Luce likelihood functions are not subject
to such a phenomenon and furnish an objective and coherent method for
ranking competitors.  Further, likelihood-based methods offer one
advantage over points-based systems: the ability to conduct
statistically rigorous tests of meaningful nulls.

@@

Table~\ref{rawresults} shows the first few and last few entries in the
dataset we use.  The first line shows that, at one of the heats in
WCT01 (Coolangatta, QLD), Colapinto came first, Bailey second, and
Wright came third: we write
$\mbox{Colapinto}\succ\mbox{Bailey}\succ\mbox{Wright}$.  The last line
shows that, at WCT11 (Banzai Pipeline), Ferreira came first and Medina
second: $\mbox{Ferreira}\succ\mbox{Medina}$.  We can convert this
dataset into a Plackett-Luce likelihood function but first have to
remove the wildcards and also player Vries whose maximum likelihood
strength is zero.  For the first line we would have Plackett-Luce
likelihood function as

@@

\noindent [observe that there is no requirement here for the strengths
  to have unit sum].  The entire dataset has a likelihood function
that incorporates all 477 lines of Table~\ref{rawresults}, but it
includes:

@@

Paired comparisons are particularly favourable to Bradley-Terry
because the likelihood function is exponential family and as such is
amenable to analysis via general linear models~\citep{turner2012}.
However, this is not applicable when more than two entities are
compared; and indeed here we see comparisons between three surfers
and, in the case of Jeep Surf Freshwater Pro WCT08, many more.  It
is natural to ask how much information is contained in the
observations where more than two surfers are in the water.

@@

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\citep{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \citep{edwards1992}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result actually reflects genuine
differences in abilities, rather than the result of random chance.
Alternatively, we may view the dataset as being sufficiently
informative to reveal actual differences in surfing ability.  Observe
that the competitors' points (that is, the final column of
table~\ref{resultstable}) cannot be used for this type of
probabilistic analysis: the points awarded are arbitrary and changing
the points system would change the points scored but not affect the
competitors' behaviour.

@@

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact that six out of eight quarterfinalists at the 2011 Nike
Lowers Pro were Brazilian; it rose in public consciousness following
the victories of Medina and de Souza in the World Surf League
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \citep{douglas2020}.

@@

\subsection{Brazilian wave type preferences}

@@

We now assess the null that Brazilians perform equally well on all
types of wave types.  To that end, we create a reified surfer,
``Brazilian wavetype", whose strength helps all Brazilian surfers
when surfing on their (putative) preferred wave type, an analogue of
the ``home ground monster" who helps the home team as per
equation~\ref{home_ground_advantage}.  Here we modify the likelihood
function only when a single Brazilian is in the water, thus avoiding
possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performance is better on their preferred wave types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
wavetype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_wavetype_support}).

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\citep{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\citep{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic team would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  Such hypotheses are difficult to test with conventional
techniques as the competitors' strengths themselves are nuisance
parameters~\citep{basu1977}.

@@

We analysed results from the 2019 World Surf League using reified
Bradley-Terry likelihoods to assess a number of hypotheses of interest
to the surfing community.  Reified Bradley-Terry allows the use of
formal statistical methods in ways that analysing points scored does
not.  For example, we can be confident that the results reflect
differences in skill (rather than random variation); there is no
statistical difference in strength between the top two officially
ranked competitors, Medina and Ferreira; but we may reject the
hypothesis that Medina and Wilson are equal in strength.

@@

\begin{table}[h]
\begin{tabular}{llllllllllllll}
competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots &WCT09&WCT10&WCT11&points\\ \hline
 Ferreira   &   BRA      &   1   &   5   &   17  & \ldots      &  2   &   1   &   1   &   59740\\
 Medina     &   BRA      &   5   &   5   &   17  & \ldots      &  9   &   9   &   2   &   56475\\
 Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots      &  9   &   2   &   17  &   49985\\
 Toledo     &   BRA      &   9   &   2   &   5   & \ldots      &  17  &   5   &   17  &   49145\\
 Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655\\
 $\ldots$   &            &   &   &       &       & \dots       &      &       &       &
\end{tabular}
\caption{Conventional league table for the top five competitors in WCT
  2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see
  Ferreira coming first, Andino second and Smith third}
\end{table}

\begin{table}[h]
\begin{verbatim}
WCT01 Colapinto Bailey Wright
WCT01 Freestone Lau Smith
WCT01 Dora Ferreira Slater
WCT01 Duru Toledo Ibelli
WCT01 Moniz Heazlewood Wilson
...
WCT11 Medina Florence
WCT11 Colapinto Bourez
WCT11 Ferreira Slater
WCT11 Medina Colapinto
WCT11 Ferreira Medina
\end{verbatim}
\caption{Extract from observations: raw results\label{rawresults} from
  WCT 2019.  Thus the first heat at WCT01 (Coolangatta, QLD) shows
  competitors Colapinto, Bailey, and Wright in the water at the same
  time, the competition order being
  $\mbox{Colapinto}\succ\mbox{Bailey}\succ\mbox{Wright}$; as discussed
  in the text we use only the competition order for our likelihood
  function}
\end{table}

@@
